---
title: "Analysis"
author: "Phani Sai Kamal Lingam"
date: "12/4/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#install.packages("units")
install.packages("ggplot2")
```

```{r}
#library("units")
library("ggplot2")
```

## Data Cleansing

```{r}
dataSet <- read.csv(file = "framingham.csv")
head(dataSet)
```

```{r}
summary(dataSet)
```


```{r}
data <- dataSet[complete.cases(dataSet), ]
head(data)
```

```{r}
summary(data)
```

```{r}
names(data) <- c("Gender", "Age", "Education", "SmokingBehavior", "CigarettesPerDay", "BloodPressureMedication", "PrevalentStroke", "PrevalentHypertension", "DiabeticCondition", "TotalCholestrol", "SystolicBloodPressure", "DiastolicBloodPressure", "BodyMassIndex", "HeartRate", "GlucoseLevel", "TenYearCoronaryHeartDisease")
```

```{r}
head(data)
```

```{r}
data$Gender[data$Gender == 0] <- "Female"
data$Gender[data$Gender == 1] <- "Male"
data$Education[data$Education == 1] <- "High School"
data$Education[data$Education == 2] <- "General Education Development"
data$Education[data$Education == 3] <- "Vocational School"
data$Education[data$Education == 4] <- "College"
data$SmokingBehavior[data$SmokingBehavior == 0] <- "Non Smoker"
data$SmokingBehavior[data$SmokingBehavior == 1] <- "Smoker"
data$BloodPressureMedication[data$BloodPressureMedication == 0] <- "Not Under BP Medication"
data$BloodPressureMedication[data$BloodPressureMedication == 1] <- "Under BP Medication"
data$PrevalentStroke[data$PrevalentStroke == 0] <- "No"
data$PrevalentStroke[data$PrevalentStroke == 1] <- "Yes"
data$PrevalentHypertension[data$PrevalentHypertension == 0] <- "No"
data$PrevalentHypertension[data$PrevalentHypertension == 1] <- "Yes"
data$DiabeticCondition[data$DiabeticCondition == 0] <- "Non Diabetic"
data$DiabeticCondition[data$DiabeticCondition == 1] <- "Diabetic"
data$TenYearCoronaryHeartDisease[data$TenYearCoronaryHeartDisease == 0] <- "Immune"
data$TenYearCoronaryHeartDisease[data$TenYearCoronaryHeartDisease == 1] <- "Vulnerable"
```

```{r}
data$Gender <- as.factor(data$Gender)
data$Education <- as.factor(data$Education)
data$SmokingBehavior <- as.factor(data$SmokingBehavior)
data$BloodPressureMedication <- as.factor(data$BloodPressureMedication)
data$PrevalentStroke <- as.factor(data$PrevalentStroke)
data$PrevalentHypertension <- as.factor(data$PrevalentHypertension)
data$DiabeticCondition <- as.factor(data$DiabeticCondition)
data$TenYearCoronaryHeartDisease <- as.factor(data$TenYearCoronaryHeartDisease)
```


```{r}
head(data)
```

```{r}
#units(data$Age) <- "years"
#units(data$TotalCholestrol) <- "mg/dL"
#units(data$SystolicBloodPressure) <- "mmHg"
#units(data$DiastolicBloodPressure) <- "mmHg"
#units(data$BodyMassIndex) <- "kg/m^2"
#units(data$GlucoseLevel) <- "mg/dL"
#head(data)
```

```{r}
summary(data)
```

## Exploratory Analysis

```{r}
ggplot(data, aes(x = Age)) + 
  geom_histogram(bins = 30, fill = "lightblue") + 
  theme_bw() + theme_classic() +
  ggtitle("Age Distribution") + ylab("Number of People")
```

```{r}
ggplot(data, aes(x = Education, fill = Education)) + 
  geom_bar() + ggtitle("Received Level of Education") +
  geom_text(stat = 'count', aes(label =..count..), vjust = -0.5)
```

```{r}
ggplot(data, aes(x = Gender, fill = Gender)) + 
  geom_bar() + 
  geom_text(stat = 'count', aes(label =..count..), vjust = -0.5) + 
  theme_bw() + theme_classic() +
  ggtitle("Gender Distribution") + ylab("Number of People")
```

```{r}
ggplot(data, aes(x = TenYearCoronaryHeartDisease)) + 
  geom_bar(aes(fill = Gender), position = 'dodge', width = 0.5, color='black') + 
  theme_bw() + theme_classic() + 
  ylab("Number of People") + ggtitle("10 Year CHD Risk Versus Gender")
```

```{r}
ggplot(data, aes(x = CigarettesPerDay)) + 
  geom_histogram(bins = 30, fill = "gray") + 
  theme_bw() + theme_classic() +
  ggtitle("Smokers - Cigarettes Per Day Distribution") + ylab("Number of People")
```

```{r}
ggplot(data, aes(x = SmokingBehavior)) + 
  geom_bar(fill = "lightgreen") + 
  geom_text(stat = 'count', aes(label =..count..), vjust = -0.5) + 
  theme_bw() + theme_classic() +
  ggtitle("Smoking Behaviour") + ylab("Number of People")
```


```{r}
ggplot(data, aes(x = TenYearCoronaryHeartDisease)) + 
  geom_bar(aes(fill = SmokingBehavior), position = 'dodge', width = 0.5, color= 'black') + 
  theme_bw() + theme_classic() + 
  ylab("Number of People") + ggtitle("Smoking Habit Versus 10 Year CHD Risk")
```

```{r}
ggplot(data, aes(x = TotalCholestrol)) + 
  geom_density(fill = "blue", alpha = 0.5) + 
  theme_minimal() + 
  ggtitle("Distribution Total Cholestrol Levels") + ylab("Number of People")
```

```{r}
ggplot(data, aes(x = TotalCholestrol)) + 
  geom_density(aes(fill = TenYearCoronaryHeartDisease), alpha = 0.4)
```

```{r}
ggplot(data, aes(x = SystolicBloodPressure)) + 
  geom_density(fill ="orange", alpha = 0.9) + 
  theme_minimal() + 
  ggtitle("Systolic BP Levels in People") + ylab("Number of People")
```

```{r}
ggplot(data, aes(x = SystolicBloodPressure)) + 
  geom_density(aes(color = TenYearCoronaryHeartDisease, fill = TenYearCoronaryHeartDisease), alpha = 0.4, position = "identity") +
  scale_fill_manual(values = c("#00AFBB", "#E7B800")) +
  scale_color_manual(values = c("#00AFBB", "#E7B800"))
```

```{r}
# Systolic Blood Pressure vs Ten Year Coronary Heart Disease
ggplot(data, aes(x = SystolicBloodPressure))+
  geom_histogram(bins = 30, color="black", fill="white")+
  facet_grid(TenYearCoronaryHeartDisease ~ .)
```

```{r}
ggplot(data, aes(x =BodyMassIndex)) + 
  geom_dotplot(color = "pink", fill = "pink", binwidth = 1/4)
```

```{r}
ggplot(data, aes(x = TenYearCoronaryHeartDisease, y = BodyMassIndex)) + 
  geom_boxplot(width = 0.4, fill = "white") +
  geom_jitter(aes(color = TenYearCoronaryHeartDisease, shape = TenYearCoronaryHeartDisease), 
              width = 0.1, size = 1) +
  scale_color_manual(values = c("#00AFBB", "#E7B800")) + 
  labs(x = NULL)
```

```{r}
ggplot(data, aes(x = TenYearCoronaryHeartDisease)) + 
  geom_bar() + 
  geom_text(stat = 'count', aes(label =..count..), vjust = -0.5) + 
  theme_bw() + theme_classic() +
  ggtitle("10 Year CHD Risk") + ylab("Number of People")
```

## Statistical Analysis

### 1. One Sample t-Test

```{r}
qqnorm(data$SystolicBloodPressure, main ="Normality Check for Systolic Blood Pressure Level")
qqline(data$SystolicBloodPressure)
```

```{r}
hist(data$SystolicBloodPressure)
```

#### (a) Traditional Statistical Tools

##### Hypothesis:

$$ H_{0}\ :\ \mu = 120\ mmHg $$

###### Null Hypothesis: The true mean Systolic Blood Pressure Level of people is 120 mmHg

$$ H_{A}\ :\ \mu \neq 120\ mmHg $$

###### Alternate Hypothesis: The true mean Systolic Blood Pressure Level of people is different than 120 mmHg

##### Parameter

###### The population parameter we want to make inference to is

$$ \mu $$

##### Sample Statistic

###### The sample statistic is the sample mean Systolic Blood Pressure

$$ \overline{x} $$

##### Test Statistic

$$ t_{n-1} = \frac{\overline{x} - \mu_0}{\frac{s}{\sqrt{n}}} $$

```{r}
# the parts of the test statistic
# sample mean
x_bar <- mean(data$SystolicBloodPressure)
# null hypothesized population mean
mu_0 <- 120
# sample st. dev
s <- sd(data$SystolicBloodPressure)
# sample size
n <- length(data$SystolicBloodPressure)
# t-test test statistic
t <- (x_bar - mu_0)/(s/sqrt(n))
t
```

##### P-Value

```{r}
# two-sided p-value so multiply by 2
two_sided_t_pval <- pt(q = t, df = n-1, lower.tail = FALSE)*2
two_sided_t_pval
```

##### Confidence Interval

```{r}
qt(0.025, n-1)
```

```{r}
# lower bound
x_bar + (qt(0.025, n-1)*(s/sqrt(n))) # alternately you can use x_bar-(qt(0.975, n-1)*(s/sqrt(n)))
```

```{r}
# upper bound
x_bar + (qt(0.975, n-1)*(s/sqrt(n))) # alternately you can use x_bar-(qt(0.025, n-1)*(s/sqrt(n)))
```

##### Interpretation

###### There is strong evidence (p-value = 5.135369e-219) to suggest that the true mean Systolic Blood Pressure Level of people is different from the given mean of 120 mmHg. We reject the null hypothesis that the true mean Systolic Blood Pressure Level of people is 26 minutes at the level. With 95% confidence, the true mean Systolic Blood Pressure Level is between 131.6546 mmHg and 133.0865 mmHg which suggests that the true mean commute time is greater than 120 mmHg.

##### R built in t.test

```{r}
t.test(data$SystolicBloodPressure, alternative = "two.sided", mu = 120)
```

#### (b) Bootstrap Methods

##### Bootstrap Approach

```{r}
set.seed(0)
# This data is pretty skewed so even though n is large, I'm going to do a lot of simulations
num_sims <- 10000
# A vector to store my results
results <- rep(NA, num_sims)
# A loop for completing the simulation
for(i in 1:num_sims){
 results[i] <- mean(sample(x = data$SystolicBloodPressure,
 size = n,
 replace = TRUE))
}
# Finally plot the results
hist(results, freq = FALSE, main='Sampling Distribution of the Sample Mean', xlab = 'Average Systolic Blood Pressure', ylab = 'Density')
# estimate a normal curve over it - this looks pretty good!
lines(x = seq(130, 134, .01), dnorm(seq(130, 134, .01), mean = x_bar, sd = s/sqrt(n)))
```

```{r}
set.seed(0)
# Shift the sample so that the null hypothesis is true
bp_given_H0_true <- data$SystolicBloodPressure - mean(data$SystolicBloodPressure) + mu_0
# This data is pretty skewed so even though n is large, I'm going to do a lot of simulations
num_sims <- 10000
# A vector to store my results
results_given_H0_true <- rep(NA, num_sims)
# A loop for completing the simulation
for(i in 1:num_sims){
 results_given_H0_true[i] <- mean(sample(x = bp_given_H0_true,
 size = n,
 replace = TRUE))
}
# Finally plot the results
hist(results_given_H0_true, freq = FALSE, main='Sampling Distribution of the Sample Mean, Given Null Hypothesis is True', xlab = 'Average Systolic Blood Pressure', ylab = 'Density')
# add line to show values more extreme on upper end
abline(v=x_bar, col = "red")
# add line to show values more extreme on lower end
low_end_extreme <- mean(results_given_H0_true)+(mean(results_given_H0_true)-x_bar)
lines(x = seq(117, 122, .01), dnorm(seq(117, 122, .01), mean = mean(results_given_H0_true), sd = sd(results_given_H0_true)))
abline(v=low_end_extreme, col="red")
```

```{r}
# counts of values more extreme than the test statistic in our original sample, given H0 is true
# two sided given the alternate hypothesis
count_of_more_extreme_lower_tail <- sum(results_given_H0_true <= low_end_extreme)
count_of_more_extreme_upper_tail <- sum(results_given_H0_true >= x_bar)
bootstrap_pvalue <- (count_of_more_extreme_lower_tail + count_of_more_extreme_upper_tail)/num_sims
bootstrap_pvalue
```

```{r}
# two sided t p-value
two_sided_t_pval
```

```{r}
# need the standard error which is the standard deviation of the results
bootstrap_SE_X_bar <- sd(results)
# an estimate is to use the formula statistic +/- 2*SE
c(x_bar - 2*bootstrap_SE_X_bar, x_bar + 2*bootstrap_SE_X_bar)
```

```{r}
# you can also use the 5th and 95th quantiles to determine the bounds:
c(quantile(results, c(.025, .975)))
```

```{r}
# compare to our t-methods
c(x_bar+(qt(0.025, n-1)*(s/sqrt(n))), x_bar+(qt(0.975, n-1)*(s/sqrt(n))))
```

### 2. One Sample Test of Proportion

#### (a) Traditional Statistical Tools

##### Hypotheses

$$ H_{0}\ :\ p_F = 0.48 $$

###### Null Hypotheses: The true proportion of Female in the population is 52% 

$$ H_{A}\ :\ p_{R} > 0.48 $$

###### Alternate Hypotheses: The true proportion of Female in the population is greater than 52% 

```{r}
p_0 <- 0.48
p_0
p <- length(data$Gender[data$Gender == "Female"])
p
n <- length(data$Gender)
n
p_hat <- p/n
p_hat
```

##### Parameter

###### The population parameter we want to make inference to is the population proportion females in the given population

$$ p_F $$

##### Sample Statistic

$$ The\ sample\ statistic\ is\ \hat{p} = \frac{2035}{3658} = 0.5563149 $$

##### Test Statistic

$$ z = \frac{p - p_0}{\sqrt{\frac{p_0 \times (1 - p_0)}{n}}}$$

```{r}
z <- (p_hat - p_0) / sqrt((p_0*(1-p_0)) / n)
z
```

##### P-Value

```{r}
binom.test(x = p, n = n, p = p_0, alternative = "greater")
```

```{r}
pnorm(z, lower.tail = FALSE)
```

##### Confidence Interval

```{r}
cat("Exact Binomial Test")
```

```{r}
binom.test(x = p, n = n, p = p_0, alternative = "greater")$conf.int
```

```{r}
cat("Normal Approx")
```

```{r}
c(p_hat - (1.64)*sqrt(((p_hat)*(1-p_hat))/n), 1)
```

##### Interpretation

###### Using the exact binomial methods for a one-sample test of proportion, there is strong evidence (p-value = 5.141e-12) to suggest that the true proportion of Female in the population is greater than 48%. We can successfully reject the null hypothesis that the true proportion of male in the population is equal to 48% at the level. The true proportion of male in the population is between 0.5428433 and 1.0000000.

#### (b) Bootstrap Methods

##### Bootstrap

```{r}
female <- data$Gender
female
```

```{r}
female <- relevel(female, "Male")
levels(female) <- c(0, 1)
female
```

```{r}
table(female)
```

```{r}
set.seed(0)
# This data is pretty skewed so even though n is large, I'm going to do a lot of simulations
num_sims <- 10000
# A vector to store my results
results <- rep(NA, num_sims)
# A loop for completing the simulation
for(i in 1:num_sims){
 results[i] <- mean(as.numeric(sample(x = female, size = n, replace = TRUE))-1)
}
# Finally plot the results
hist(results, freq = FALSE, main='Sampling Distribution of the Sample Proportion', xlab = 'Proportion of Female', ylab = 'Density')
# estimate a normal curve over it - this looks pretty good!
lines(x = seq(.52, .60, .001), dnorm(seq(.52, .60, .001), mean = mean(results), sd = sd(results)))
```

```{r}
cat("Bootstrap Confidence Interval")
```

```{r}
c(quantile(results, c(0.05,1)))
```

```{r}
cat("exact binomial test")
```

```{r}
binom.test(x = p, n = n, p = p_0, alternative = "greater")$conf.int
```

```{r}
cat("normal approx")
```

```{r}
c(p_hat - (1.64)*sqrt(((p_hat)*(1-p_hat))/n),1)
```

```{r}
# Under the assumption that the null hypothesis is true, we have 48% female
female_sim <- rep(c(1, 0), c(.48*n, (1-.48)*n))
num_sims <- 10000
# A vector to store my results
results_H0_true <- rep(NA, num_sims)
# A loop for completing the simulation
for(i in 1:num_sims){
 results_H0_true[i] <- mean(sample(x = female_sim,
 size = n,
 replace = TRUE))
}
# Finally plot the results
hist(results_H0_true, freq = FALSE, main='Sampling Distribution of the Sample Proportion under H_0:p = 0.48', xlab = 'Proportion of Female', ylab = 'Density')
# estimate a normal curve over it - this looks pretty good!
lines(x = seq(.30, .65, .001), dnorm(seq(.30, .65, .001), mean = mean(results_H0_true), sd = sd
(results_H0_true)))
abline(v=p_hat, col="red")
```

```{r}
count_of_more_extreme_upper_tail <- sum(results_H0_true >= p_hat)
bootstrap_pvalue <- count_of_more_extreme_upper_tail/num_sims
cat("Bootstrap p-value")
```

```{r}
bootstrap_pvalue
```

```{r}
cat("Exact Binomial p-value")
```

```{r}
binom.test(x = p, n = n, p = p_0, alternative = "greater")$p.value
```

```{r}
cat("Normal Approximation p-value")
```

```{r}
pnorm(z, lower.tail = FALSE)
```

### 3. Two sample t-Test for Difference in Means

```{r}
qqnorm(data$SystolicBloodPressure, main ="Normality Check for Systolic Blood Pressure Level")
qqline(data$SystolicBloodPressure)
```

```{r}
qqnorm(data$SystolicBloodPressure[data$TenYearCoronaryHeartDisease == "Vulnerable"], main ="Normality Check for Systolic Blood Pressure of patients Vulnerable to 10 Year CHD")
qqline(data$SystolicBloodPressure[data$TenYearCoronaryHeartDisease == "Vulnerable"])
```

```{r}
qqnorm(data$SystolicBloodPressure[data$TenYearCoronaryHeartDisease == "Immune"], main ="Normality Check for Systolic Blood Pressure of patients Immune to 10 Year CHD")
qqline(data$SystolicBloodPressure[data$TenYearCoronaryHeartDisease == "Immune"])
```

#### Sub-Sampling

```{r}
set.seed(0)
immunePatientsData <- subset(data, data$TenYearCoronaryHeartDisease == "Immune")
immuneDataSample <- immunePatientsData[sample(nrow(immunePatientsData), 300), ]
head(immuneDataSample)
```

```{r}
set.seed(0)
vulnerablePatientsData <- subset(data, data$TenYearCoronaryHeartDisease == "Vulnerable")
vulnerableDataSample <- vulnerablePatientsData[sample(nrow(vulnerablePatientsData), 300), ]
head(vulnerableDataSample)
```

```{r}
sampleData <- rbind(immuneDataSample, vulnerableDataSample)
head(sampleData)
```

```{r}
summary(sampleData)
```

```{r}
qqnorm(sampleData$SystolicBloodPressure, main ="Sample Data - Normality Check for Systolic Blood Pressure Level")
qqline(sampleData$SystolicBloodPressure)
```

```{r}
qqnorm(sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease == "Vulnerable"], main ="Sample - Data Normality Check for Systolic Blood Pressure of patients Vulnerable to 10 Year CHD")
qqline(sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease == "Vulnerable"])
```

```{r}
qqnorm(sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease == "Immune"], main ="Sample Data - Normality Check for Systolic Blood Pressure of patients Immune to 10 Year CHD")
qqline(sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease == "Immune"])
```

#### (a) Traditional Statistical Tools

##### Hypothesis

###### Null Hypothesis: The true population mean Systolic Blood Pressure of Patients Vulnerable to Heart Disease is equal to the true population mean Systolic Blood Pressure of Patients Immune to Heart Disease.

$$ H_0: \mu_v - \mu_i = 0\ or\ \mu_v = \mu_i $$

###### Alternate Hypothesis: The true population mean Systolic Blood Pressure of Patients Vulnerable to Heart Disease is not equal to the true population mean Systolic Blood Pressure of Patients Immune to Heart Disease.

$$ H_{A1}: \mu_v - \mu_i \ne 0\ or\ \mu_v \ne \mu_i$$

##### Parameter

###### We are interested in the true population mean difference in Systolic Blood Pressure Levels between those who are Vulnerable to Heart Disease and those who are Immune to Heart Disease

$$ \overline{\mu}_v - \overline{\mu}_i$$

##### Sample Statistic: Difference in Means

$$ \overline{x}_v - \overline{x}_i$$

##### Test-statistic 

$$ t = \frac{(\overline{x}_v - \overline{x}_i) - (\mu_v - \mu_i)}{\sqrt{\frac{\sigma^2_v}{n_v} + {\frac{\sigma^2_i}{n_i}}}} $$

$$ \mu_0 = \mu_v - \mu_i = 0 \\ $$

##### Two Sample t-test

###### Calculating p-value using R in-built function t.test()

```{r}
t.test(sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease == "Vulnerable"], sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease == "Immune"])
```

###### Calculating p-value manually using given formulas.

```{r}
# Mean Systolic Blood Pressure of Vulnerable Patients
mu_v <- mean(sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease == 'Vulnerable'])
mu_v
```

```{r}
# Mean Systolic Blood Pressure of Immune Patients
mu_i <- mean(sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease == 'Immune'])
mu_i
```

```{r}
# Null Hypothesis
mu_0 <- 0
```

```{r}
# Variance of Systolic Blood Pressure of Vulnerable Patients
var_v <- var(sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease == 'Vulnerable'])
var_v
```

```{r}
# Variance of Systolic Blood Pressure of Vulnerable Patients
var_i <- var(sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease == 'Immune'])
var_i
```

```{r}
# Sample Size of Systolic Blood Pressure of Vulnerable Patients
n_v <- length(sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease == 'Vulnerable'])
n_v
```

```{r}
# Sample Size of Systolic Blood Pressure of Vulnerable Patients
n_i <- length(sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease == 'Immune'])
n_i
```

```{r}
# t-value (test statistic)
t <- (mu_v - mu_i - mu_0)/sqrt(var_v/n_v + var_i/n_i)
t
```

```{r}
# p-value for 2 sided t-test
p_value <- pt(q = t, df = min(n_v, n_i) - 1, lower.tail = FALSE)*2
p_value
```

```{r}
# Lower Boundary of Confidence Interval
lowerBound <- mu_v - mu_i + qt(0.05, min(n_v, n_i) - 1)*sqrt(var_v/n_v + var_i/n_i)
lowerBound
```

```{r}
# Upper Boundary of Confidence Interval
upperBound <- mu_v - mu_i + qt(0.95, min(n_v, n_i) - 1)*sqrt(var_v/n_v + var_i/n_i)
upperBound
```

#### (b) Bootstrap Methods

```{r}
set.seed(0)
num_sims <- 10000
# A vector to store my results
results <- rep(NA, num_sims)
# A loop for completing the simulation
for(i in 1:num_sims){
 mean_immune <- mean(sample(x = sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease == 'Immune'],
 size = 300,
 replace = TRUE))
 mean_vulnerable <- mean(sample(x = sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease == 'Vulnerable'],
 size = 300,
 replace = TRUE))
 results[i] <- mean_vulnerable - mean_immune
}
# Finally plot the results
hist(results, freq = FALSE, main='Sampling Distribution of the Sample Mean', xlab = 'Average Difference Systolic Blood Pressure', ylab = 'Density')
lines(x = seq(9, 21, .01), dnorm(seq(9, 21, .01), mean = mean(results), sd = sd(results)))
```

```{r}
# Bootstrap one-sided CI
c(quantile(results, c(.025, .975)))
```

```{r}
t.test(sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease == "Vulnerable"], sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease == "Immune"])$conf.int
```

```{r}
set.seed(0)
transform(sampleData,Group=sample(TenYearCoronaryHeartDisease))
```

```{r}
set.seed(0)
num_sims <- 10000
# A vector to store my results
results_given_H0_true <- rep(NA, num_sims)
# A loop for completing the simulation
for(i in 1:num_sims){
 # idea here is if there is no relationshipm we should be able to shuffle the groups
  shuffled_groups <- transform(sampleData,Group=sample(TenYearCoronaryHeartDisease))
  mean_immune <- mean(shuffled_groups$SystolicBloodPressure[shuffled_groups$Group=="Immune"])
  mean_vulnerable <- mean(shuffled_groups$SystolicBloodPressure[shuffled_groups$Group=="Vulnerable"])
  results_given_H0_true[i] <- mean_vulnerable - mean_immune
}
# Finally plot the results
hist(results_given_H0_true, freq = FALSE, main='Dist. of the Diff in Sample Means Under Null', xlab = 'Average Difference Systolic Blood Pressure under Null', ylab = 'Density')
diff_in_sample_means <- mean(sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease=="Vulnerable"]) - mean(sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease=="Immune"])
low_end_extreme <- mean(results_given_H0_true) + (mean(results_given_H0_true) - diff_in_sample_means)
lines(x = seq(-6, 6, .01), dnorm(seq(-6, 6, .01), mean = mean(results_given_H0_true), sd = sd(results_given_H0_true)))
abline(v=diff_in_sample_means, col = "blue")
abline(v=abs(diff_in_sample_means), col = "red")
```

```{r}
# counts of values more extreme than the test statistic in our original sample, given H0 is true
# two sided given the alternate hypothesis
count_of_more_extreme_lower_tail <- sum(results_given_H0_true <= low_end_extreme)
count_of_more_extreme_upper_tail <- sum(results_given_H0_true >= diff_in_sample_means)
bootstrap_pvalue <- (count_of_more_extreme_lower_tail + count_of_more_extreme_upper_tail)/num_sims
cat("Bootstrap p-value")
```

```{r}
bootstrap_pvalue
```

```{r}
cat("t-test p-value")
```

```{r}
t.test(sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease == "Vulnerable"], sampleData$SystolicBloodPressure[sampleData$TenYearCoronaryHeartDisease == "Immune"])$p.value
```

##### Interpretation

###### There is strong evidence (p-value=0.056) to suggest that the true population mean Systolic Blood Pressure for the Patients Vulnerable to Heart Disease is different than those who are Immune too Heart Disease. We succeed to reject the null hypothesis that there is no difference between the mean Systolic Blood Pressure between the Vulnerable and Immune groups at the level. With 95% confidence, the true difference in the mean Systolic Blood Pressure between those who are Vulnerable and those who are Immune is between 11.47974 and 18.99359. The null hypothesized difference between the mean Systolic Blood Pressure is zero and zero is not in the 95% confidence interval - this result is not consistent with the results of our hypothesis test but it is possible to have this type of inconsistency when using the bootstrap methods. The values of the confidence interval suggest that on average those who are Immune to Heart Disease have a lower Systolic Blood Pressure Level than those who are Vulnerable to Heart Diesease.

### 4. Two sample test for Difference in Proportions

#### (a) Traditional Statistical Tools

##### Hypotheses

$$ H_0\ :\ p_M - p_F = 0 $$

###### Null Hypotheses: There is no difference between the true population proportion of Male Patients Vulnerable to Heart Disease and the true population proportion of Female Patients Vulnerable to Heart Disease.

$$ H_A\ :\ p_M - p_F \ne 0 $$

###### Null Hypotheses: There is a difference between the true population proportion of Male Patients Vulnerable to Heart Disease and the true population proportion of Female Patients Vulnerable to Heart Disease.

##### Paramter

###### We are in interested in the difference between the true population proportion of Male who are Vulnerable to Heart Disease and true population proportion of Female who are Vulnerable to Heart Disease.

$$ p_M - p_F$$

##### Sample Statistic

$$ \hat{p_M} - \hat{p_F} $$

##### Test Statistic

$$ z = \frac{(\hat{p_M} - \hat{p_F}) - (p_M - p_F)}{\sqrt{\frac{\hat{p_M}(1 - \hat{p_M})}{n_M}+\frac{\hat{p_F}(1 - \hat{p_F})}{n_F}}} $$

```{r}
# the parts of the test statistic
# sample props
p_hat_M <- length(data$Gender[data$Gender == "Male" & data$TenYearCoronaryHeartDisease == "Vulnerable"])/length(data$Gender[data$Gender == "Male"])
p_hat_F <- length(data$Gender[data$Gender == "Female" & data$TenYearCoronaryHeartDisease == "Vulnerable"])/length(data$Gender[data$Gender == "Female"])
# null hypothesized population prop difference between the two groups
p_0 <- 0
# sample size
n_M <- length(data$Gender[data$Gender == "Male"])
n_F <- length(data$Gender[data$Gender == "Female"])
# sample variances
den_p_M <- (p_hat_M*(1-p_hat_M))/n_M
den_p_F <- (p_hat_F*(1-p_hat_F))/n_F
# z-test test statistic
z <- (p_hat_M - p_hat_F - p_0)/sqrt(den_p_M + den_p_F)
z
# two sided p-value
two_sided_diff_prop_pval <- pnorm(q = z, lower.tail = FALSE)*2
two_sided_diff_prop_pval
```

##### Confidence interval

```{r}
# lower bound
(p_hat_M - p_hat_F)+(qnorm(0.025)*sqrt(den_p_M + den_p_F))
```

```{r}
# upper bound
(p_hat_M - p_hat_F)+(qnorm(0.975)*sqrt(den_p_M + den_p_F))
```

#### (b) Bootstrap Methods

##### Bootstrap + Randomization Approach

```{r}
# Make the data
male <- rep(c(1, 0), c(length(data$Gender[data$Gender == "Male" & data$TenYearCoronaryHeartDisease == "Vulnerable"]), n_M - length(data$Gender[data$Gender == "Male" & data$TenYearCoronaryHeartDisease == "Vulnerable"])))
female <- rep(c(1,0), c(length(data$Gender[data$Gender == "Female" & data$TenYearCoronaryHeartDisease == "Vulnerable"]), n_F - length(data$Gender[data$Gender == "Female" & data$TenYearCoronaryHeartDisease == "Vulnerable"])))
num_sims <- 10000
# A vector to store my results
results <- rep(NA, num_sims)
# A loop for completing the simulation
for(i in 1:num_sims){
 prop_M <- mean(sample(male,
 size = n_M,
 replace = TRUE))
 prop_F <- mean(sample(x = female,
 size = n_F,
 replace = TRUE))
 results[i] <- prop_M - prop_F
}
# Finally plot the results
hist(results, freq = FALSE, main='Dist. of the Diff in Prop', xlab = 'Difference in Prop. of Patients Vulnerable to Heart Disease', ylab = 'Density')
lines(x = seq(0.01, 0.13, .001), dnorm(seq(0.01, 0.13, .001), mean = mean(results), sd = sd(results)))
```

```{r}
cat("Bootstrap")
```

```{r}
c(quantile(results, c(.025, .975)))
```

```{r}
cat("Normal Approximation")
```

```{r}
c((p_hat_M - p_hat_F)+(qnorm(0.025)*sqrt(den_p_M + den_p_F)), (p_hat_M - p_hat_F)+(qnorm(0.975)*sqrt(den_p_M + den_p_F)))
```

```{r}
# Make the data
df_combined <- data.frame("vulnerable_patients" = c(male, female), "gender" = rep(c("male", "female"), c(n_M, n_F)))
# Sanity checks
summary(df_combined$gender)
```

```{r}
mean(df_combined$vulnerable_patients[df_combined$gender=="male"]) == p_hat_M
```

```{r}
mean(df_combined$vulnerable_patients[df_combined$gender=="female"]) == p_hat_F
```

```{r}
num_sims <- 1000
# A vector to store my results
results_given_H0_true <- rep(NA, num_sims)
# A loop for completing the simulation
for(i in 1:num_sims){
 # idea here is if there is no relationshipm we should be able to shuffle the groups
 shuffled_groups <- transform(df_combined, gender=sample(gender))
 prop_M <- mean(shuffled_groups$vulnerable_patients[shuffled_groups$gender=="male"
])
 prop_F <- mean(shuffled_groups$vulnerable_patients[shuffled_groups$gender=="female"
])
 results_given_H0_true[i] <- prop_M - prop_F
}
# Finally plot the results
hist(results_given_H0_true, freq = FALSE,
 main='Dist. of the Diff in Sample Sample Props Under Null',
 xlab = 'Average Difference in Prop. Vulnerable Patients under Null',
 ylab = 'Density')
diff_in_sample_props <- p_hat_M - p_hat_F
lines(x = seq(-0.05, 0.05, .001), dnorm(seq(-0.05, 0.05, .001), mean = mean(results_given_H0_true), sd = sd(results_given_H0_true)))
abline(v=diff_in_sample_props, col = "blue")
abline(v=-diff_in_sample_props, col = "red")
```

```{r}
# counts of values more extreme than the test statistic in our original sample, given H0 is true
# two sided given the alternate hypothesis
count_of_more_extreme_lower_tail <- sum(results_given_H0_true <= -diff_in_sample_props)
count_of_more_extreme_upper_tail <- sum(results_given_H0_true >= diff_in_sample_props)
bootstrap_pvalue <- (count_of_more_extreme_lower_tail + count_of_more_extreme_upper_tail)/num_sims
cat("Bootstrap p-value")
```

```{r}
bootstrap_pvalue
```

```{r}
cat("Normal Approx p-value")
```

```{r}
two_sided_diff_prop_pval
```

##### Interpretation

###### Using randomization methods, there is strong evidence (p-value = 0) to suggest that there is a difference between the true proportion of Male Vulnerable to Heart Disease compared to their Female Counterparts. We successfully reject the null hypothesis that the true proportion of fMale Vulnerable to Heart Disease is equal to the true proportion of Female Vulnerable to Heart Disease at the level. Using confidence intervals created by the bootstrap method, we can say with 95% confidence that the true population proportion difference lies between 4.2% to 9.0% which means Male are more vulnerable to heart disease than the Female. The null hypothesized difference of 0 is outside the confidence interval which agrees with our rejection of the null hypothesis.

### 5.Chi-square goodness of fit

```{r}
chiData <- data$Education
head(chiData)
```

```{r}
table(chiData)
```

```{r}
prop.table(table(chiData))
```

#### (a) Traditional Statistical Tools

##### Hypotheses

$$ H_0\ :\ p_C = p_{GED} = p_{HS} = p_{VS} = 0.25$$

###### Null Hyptheses: The proportion of each level of education is the same and is equal to 0.25

$$ H_A\ :\ Some\ p_i \neq 0.25$$

###### Alternate Hyptheses: At least one of the proportions is not equal to 0.25

##### Parameter

###### We are in interested in the true proportions of people in each level of education

$$ p_C,\ p_{GED},\ p_{HS},\ p_{VS}$$

##### Sample Statistics

$$ \hat{p_C},\ \hat{p_{GED}},\ \hat{p_{HS}},\ \hat{p_{VS}} $$

##### Test Statistic and Distribution

$$ \chi^2=\sum_{i=1}^{k} \frac{(O_i - E)^2}{E} \tilde{} \chi^2_{k-1} $$

$$ n = 3658 \\
p_i = 0.25 \\
expected\ count,\ np_i = 3658 \times 0.25 \\
np_i = 914.5 $$

```{r}
n <- 3658
r <- 4
npi <- 914.5
tchi <- sum(((table(chiData) - npi)^2)/npi)
tchi
```

##### P-Value

```{r}
p_value <- pchisq(tchi, df = r-1, lower.tail = FALSE)
p_value
```

##### Confidence interval

###### There is no confidence interval for a goodness of fit test.

#### (b) Randomization Approach

```{r}
# Create our data under the assumption that H_0 is true
solutions_under_H_0 <- rep(c("C", "GED", "HS", "VS"), npi)
# Sanity Check
table(solutions_under_H_0)
```

```{r}
num_sims <- 10000
# A vector to store my results
chisq_stats_under_H0 <- rep(NA, num_sims)
# A loop for completing the simulation
for(i in 1:num_sims){
 new_samp <- sample(solutions_under_H_0, n, replace = T)
 chisq_stats_under_H0[i] <- sum(((table(new_samp) - npi)^2)/npi)
}
```

```{r}
hist(chisq_stats_under_H0, freq = FALSE,
 main='Dist. of the Chi-Square Statistic Under Null',
 xlab = 'Chi-Square Stat under Null',
 ylab = 'Density')
abline(v=sum(((table(chiData) - npi)^2)/npi), col="red")
```

```{r}
#The randomization p-value
sum(chisq_stats_under_H0 >= sum(((table(chiData) - npi)^2)/npi))/num_sims
```

### Appendix

```{r ref.label=knitr::all_labels(), echo = T, eval = F}
```

